Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2017-11-26 paper/paper.tex
Fixing most figures.

To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of one of the waveform parameters $S_1$. The fact that the residuals are smooth functions of the waveform parameters indicates that most numerical noise has been removed. The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.05; \bx)$, where the amplitude is very small and dominated by numerical noise.

@@

For the problem here, we have found that global basis functions do not work well. The residuals are small and smooth at low frequencies and large and noisy at high frequencies. With global basis functions, errors in evaluating $g(MF_j;\bx)$ at high frequencies can propagate to large errors between the nodes $MF_j$ at low frequencies. Instead, we find that spline interpolation works significantly better. We use 40 frequency nodes $MF_j$ log-spaced in the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}]$ (see Fig.~\ref{fig:hoff}). We evaluate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at these nodes as discussed below. We then interpolate between these frequencies using third-order splines. The local, third-order polynomials, that are only connected by the requirement of smoothness do not propagate high-frequency errors down to low-frequency errors as significantly as do global basis functions.

@@

Next we choose a method to interpolate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at each of the frequency nodes $MF_j$ as a function of the five waveform parameters $\bx$. Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality; the number of samples grows exponentially with the dimension $d$ ($N^d$ samples for $N$ points per dimension). For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter. At $\sim10$~minutes--1~day per EOB waveform on a standard CPU, this is at the limit of what is reasonable. However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below. The method we choose is Gaussian Process Regression (GPR).

@@

In a zero-mean Gaussian process, the data $\by_i$ and function value $\by_*$ at the training set points $\bx_i$ and new point $\bx_*$, respectively, are drawn from a multivariate normal distribution
\begin{equation}
\label{eq:gaussian}
\begin{bmatrix}
{\bm y} \\
y_* \\
\end{bmatrix}
\sim \mathcal{N}
\left({\bm 0}, 
\begin{bmatrix}
K & K_*^T \\
K_* & K_{**} \\
\end{bmatrix}
\right)
\end{equation}
where $K_{ij} = k(\bx_i, \bx_j)$ is a matrix, $K_{*i} = k(\bx_i, \bx_*)$ is a vector, and $K_{**} = k(\bx_*, \bx_*)$ is a scalar.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{uncertaintysampling.png}
\caption{Estimated RMS phase error $\epsilon_{\rm RMS}$ (Eq.~\eqref{eq:errorrms}) maximized over parameter space $\bx$ as a function of the number of chosen parameters (blue curve). $\epsilon_{\rm RMS}$ depends on the chose samples $\bx_j$ and hyperparameters, not on the data, and can be calculated before generating the waveforms. Also shown for comparison are the maximum RMS phase errors at the nodes where $MF_j \le 0.03$ for the surrogate compared to a test set of 1000 waveforms. Blue circle: surrogate constructed from the 160 waveforms in the initial training set. Red square: surrogate after adding the 200 waveforms chosen with uncertainty sampling. Green triangle: final surrogate after also adding 1000 more training set waveforms drawn uniformly from parameter space. }
\label{fig:uncsamp}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{mismatch.png}
\caption{Mismatches between the surrogate and 1000 test-set waveforms with randomly sampled parameters $\bx$. The total mass for each waveform is $M=2.8M_\odot$ and the PSD used was the aLIGO design sensitivity PSD. Histograms are shown for the initial surrogate with 160 waveforms (blue), the updated surrogate with 360 waveforms (red), and the final surrogate with 1360 waveforms (green). Also shown for comparison are the mismatches between TaylorF2 and the test set (black).}
\label{fig:mismatch}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{htildemaxerror.png}
\caption{Amplitude and phase errors between the surrogate and the test-set waveform with the largest mismatch. Blue curve: initial surrogate with 160 waveforms. Red curve: updated surrogate with 360 waveforms. Green curve: final surrogate with 1360 waveforms. The vertical solid line is the gravitational-wave frequency at the Schwarzschild ISCO.}
\label{fig:maxmismatch}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{hmaxerror.png}
\caption{Inverse Fourier transformed surrogate (green dashed) for the parameters $\bx$ that had the largest mismatch with the test-set EOB waveforms (black). The two waveforms are aligned in the time domain in the interval $-2\times 10^5 < t/M < -1\times 10^5$.}
\label{fig:maxmismatchtd}
\end{figure}

>>>>

To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of one of the waveform parameters $S_1$. The fact that the residuals are smooth functions of the waveform parameters indicates that most numerical noise has been removed. The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.052; \bx)$, where the amplitude is very small and dominated by numerical noise.

@@

For the problem here, we have found that global basis functions do not work well. The residuals are small and smooth at low frequencies and large and noisy at high frequencies. With global basis functions, errors in evaluating $g(MF_j;\bx)$ at high frequencies can propagate to large errors between the nodes $MF_j$ at low frequencies. Instead, we find that spline interpolation works significantly better. We use 40 frequency nodes $MF_j$ log-spaced in the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}]$ (see Fig.~\ref{fig:hoff}). We evaluate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at these nodes as discussed below. We then interpolate between these frequencies using third-order splines. The local, third-order polynomials, that are only connected by the requirement of smoothness, do not propagate high-frequency errors down to low-frequency errors as significantly as do global basis functions.

@@

Next we choose a method to interpolate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at each of the frequency nodes $MF_j$ as a function of the five waveform parameters $\bx$. Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality: the number of samples grows exponentially with the dimension $d$ ($N^d$ samples for $N$ points per dimension). For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter. At $\sim10$~minutes--1~day per EOB waveform on a standard CPU, this is at the limit of what is reasonable. However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below. The method we choose is Gaussian Process Regression (GPR).

@@

In a zero-mean Gaussian process, the data $y_i$ and function value $y_*$ at the training set points $\bx_i$ and new point $\bx_*$, respectively, are drawn from a multivariate normal distribution
\begin{equation}
\label{eq:gaussian}
\begin{bmatrix}
{\bm y} \\
y_* \\
\end{bmatrix}
\sim \mathcal{N}
\left({\bm 0}, 
\begin{bmatrix}
K & K_*^T \\
K_* & K_{**} \\
\end{bmatrix}
\right)
\end{equation}
where $K_{ij} = k(\bx_i, \bx_j)$ is a matrix, $K_{*i} = k(\bx_i, \bx_*)$ is a vector, and $K_{**} = k(\bx_*, \bx_*)$ is a scalar.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{uncertaintysampling.pdf}
\caption{Estimated RMS phase error $\epsilon_{\rm RMS}$ (Eq.~\eqref{eq:errorrms}) maximized over parameter space $\bx$ as a function of the number of chosen parameters (blue curve). $\epsilon_{\rm RMS}$ depends on the chose samples $\bx_j$ and hyperparameters, not on the data, and can be calculated before generating the waveforms. Also shown for comparison are the maximum RMS phase errors at the nodes where $MF_j \le 0.03$ for the surrogate compared to a test set of 1000 waveforms. Blue circle: surrogate constructed from the 160 waveforms in the initial training set. Red square: surrogate after adding the 200 waveforms chosen with uncertainty sampling. Green triangle: final surrogate after also adding 1000 more training set waveforms drawn uniformly from parameter space. }
\label{fig:uncsamp}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{mismatch.pdf}
\caption{Mismatches between the surrogate and 1000 test-set waveforms with randomly sampled parameters $\bx$. The total mass for each waveform is $M=2.8M_\odot$ and the PSD used was the aLIGO design sensitivity PSD. Histograms are shown for the initial surrogate with 160 waveforms (blue), the updated surrogate with 360 waveforms (red), and the final surrogate with 1360 waveforms (green). Also shown for comparison are the mismatches between TaylorF2 and the test set (black).}
\label{fig:mismatch}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{htildemaxerror.pdf}
\caption{Amplitude and phase errors between the surrogate and the test-set waveform with the largest mismatch. Blue curve: initial surrogate with 160 waveforms. Red curve: updated surrogate with 360 waveforms. Green curve: final surrogate with 1360 waveforms. The vertical solid line is the gravitational-wave frequency at the Schwarzschild ISCO.}
\label{fig:maxmismatch}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{hmaxerror.pdf}
\caption{Inverse Fourier transformed surrogate (green dashed) for the parameters $\bx$ that had the largest mismatch with the test-set EOB waveforms (black). The two waveforms are aligned in the time domain in the interval $-2\times 10^5 < t/M < -1\times 10^5$.}
\label{fig:maxmismatchtd}
\end{figure}
